---
title: "White wine quality"
output: html_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
#Introduction
In this project i choose to work with **white wine quality** dataset. White wine
is a wine whose color can be straw-yellow, yellow-green or yellow-gold.It is 
produced by the alcoholic fermentation of the non-coloured pulp of the grapes,
which may have a skin of any color. It is very difficult to assess the quality 
of wine by reading label. Quality is assessed best by tasting. Quite often you 
won't be able to taste a wine before buying, but information is available on 
wine lables to help you decide, although you will need a certain amount of the
knowledge about the contents of the wine in prior.
```{r echo=FALSE, message= FALSE,warning=FALSE,packages}
library(ggplot2)
library(gridExtra)
```

```{r echo=FALSE,message=FALSE,warning=FALSE,loading_data}
#load the white wine quality dataset
wine <- read.csv('wineQualityWhites.csv')
# remove the first coloumn x
wine$X <- NULL
```

#Univariate Plot Section
```{r echo=FALSE,message=FALSE,warning=FALSE,summary}
dim(wine)
str(wine)
```
There are 4898 observation and 12 varaiables to denote the composition and 
quality of white wine. From 12 of the variable 11 of them are numeric datatype 
and an integer to denote the quality of wine.

__alcohol__

```{r echo=FALSE,message=FALSE,warning=FALSE,plotting}
ggplot(aes(x = alcohol),data = wine)+
  scale_x_continuous(breaks = seq(8,15,0.4))+
  geom_histogram(binwidth = 0.2,color = I('black'),fill = I('white'))
summary(wine$alcohol)
```

In normal most of the white wine can have a alcohol content ranging from 5%-15%.
In this dataset the alcohol content ranging from 8% to 14.20% and 50% of the 
wines are below 10.40% alcohol content.

__residual sugar__

```{r echo=FALSE,message=FALSE,warning=FALSE,residualsugar}
wine$taste_type <- cut(wine$residual.sugar,breaks = c(0.0,1.01,10.01,120),
                       labels = c('bone_dry','dry','off-dry'))
wine$taste_type <- factor(wine$taste_type)
ggplot(aes(x = residual.sugar),data = wine)+
  scale_x_continuous(limits = c(0,20),breaks = seq(0,20,1))+
  geom_histogram()
summary(wine$residual.sugar)
```
Residual sugars are the sugars that are not converted into alcohol during 
fermentation. Residual sugar in wines ranges from 0 to 220 grams per liter sugar
depanding on the style. Here it ranges from 0.600 to 65.800 with half of the 
wines fall below 5.200 and 75% of the wines have less than 9.900 residual sugar.
Less than 25% wines have residual sugar greater than 10.0.

```{r echo=FALSE,warning=FALSE,message=FALSE,residual.sugar.content}
ggplot(aes(x=residual.sugar),data=subset(wine,wine$residual.sugar >= 10.0 &
                                             wine$residual.sugar <= 40 ))+
  geom_histogram(fill='blue')+
  ggtitle("residual sugar content between 10 and 40")
```

From the 25% wines that are below 9.900 most of them fall in the range of 10-20.
Only few wines have residual sugar greater than 20.

**Taste Type**

Based on the residual sugar content in the wine , wines are classified into 
three type.

```{r echo=FALSE,message=FALSE,warning=FALSE,taste_type}
ggplot(aes(x=taste_type),data=wine)+
  geom_histogram(stat = "count")
table(wine$taste_type)
```

Most of the wines are dry type wine.There is no wide spread sugar content in 
our dataset.

**volatile acidity**

In US legal limit of the volatile acidity is 1.1g/L. When volatile acidity 
becomes higher it gives the smell and taste of vinegar to the wine.

```{r echo=FALSE,message=FALSE,warning=FALSE,volatile_acidity}
ggplot(aes(x = volatile.acidity),data = wine)+
  scale_x_continuous(limits = c(0.0,1.1),breaks = seq(0,1.1,0.1))+
  geom_histogram( binwidth = 0.1)
summary(wine$volatile.acidity)

```
About 75% of the wine has volatile acidity less than .3200.There is a steady 
decrease in volatile acidity after 0.3.

**fixed acidity**

```{r echo=FALSE,warning=FALSE,message=FALSE,fixedacidity}
ggplot(aes(x = fixed.acidity),data = wine)+
  geom_histogram()
summary(wine$fixed.acidity)
```

Most of the data falls between the range of 6.3 to 7.3

**PH**

winemakers uses pH as a way to measure the acidity of the wine.Low pH wines will
taste tart or crisp while the high pH wine are more susceptible for bacterial 
growth.Most of the wine pH fall around 3 to 4. pH of 3.0 to 3.4 is suitable for
white wines.

```{r echo = FALSE,warning = FALSE,message = FALSE,PH}
pH1 <- ggplot(aes(x = pH),data = wine)+
  geom_histogram(color = 'black',fill = 'white')+
  ggtitle('pH of all the white wines')
pH2 <- ggplot(aes(x = pH),data = wine)+
  scale_x_continuous(limits = c(3.0,3.4),breaks = seq(3.0,3.4,0.05))+
  geom_histogram(color = 'black',fill = 'white',binwidth = 0.025)+
  ggtitle('pH between 3 to 3.4')
pH3 <- ggplot(aes(x = pH),data = wine)+
  scale_x_continuous(limits = c(3.4,3.8),breaks = seq(3.4,3.8,0.05))+
  geom_histogram(color = 'black',fill = 'white',binwidth = 0.025)+
  ggtitle('pH between 3.4 to 3.8')
grid.arrange(pH1,pH2,pH3)
summary(wine$pH)
```

Here most of the wines have the pH within the range of 3 to 3.4.There are very
few wines below 3.6pH.

**citric acid**

citric acid is minor in grapes, added after fermentation to add flavours to wine
```{r echo=FALSE,message=FALSE,warning=FALSE,citric.acid}
ggplot(aes(x = citric.acid), data = wine)+
  geom_histogram()
summary(wine$citric.acid)
```

Most of the wine has less than 0.5g/L .There are also some wines with 0 citric
acid in their content. when wine have low acidity then citric acid is added to 
increase their flavours to them.

**chlorides**

chloride content in wine is due to the vineyard which is present near sea or the
irrigation by the salt water. In most cases the limit is 60mg/L or 0.06g/L.

```{r echo=FALSE,message=FALSE,warning=FALSE,chlorides}
c1 <- ggplot(aes(x = chlorides),data = wine)+
  xlim(0,quantile(wine$chlorides,0.95))+
  geom_histogram(binwidth = 0.01)
c2 <- ggplot(aes(x = chlorides),data = wine)+
  scale_x_continuous(limits = c(0.06,0.3),breaks = seq(0.06,0.3,0.02))+
  geom_histogram(binwidth = 0.02)
grid.arrange(c1,c2)
summary(wine$chlorides)
```

More than 75% of the wines have chlorides less than 0.06. There are also some
wines higher than 0.06 but their count is very low.

**sulfur dioxide **

sulphates are added to produce sulfur dioxide in wine so that it can prevent it 
from oxidation and spoilage.sulfur dioxide is present in wine as free SO2 and 
also bound to acetaldehyde.The sum gives the total sulfur dioxide in wine. Legal
limit of SO2 is 350mg/L and the sweet wine is 400mg/L.

```{r echo=FALSE,warning=FALSE,message=FALSE,sulfurdioxide}
wine <- transform(wine,bound.sulfur.dioxide = total.sulfur.dioxide - 
                    free.sulfur.dioxide)
p1 <- ggplot(aes(x = free.sulfur.dioxide),data = wine)+
  xlim(0,quantile(wine$free.sulfur.dioxide,0.95))+
  geom_histogram(color = 'black',fill = 'white')+
  geom_vline(aes(xintercept = median(free.sulfur.dioxide)),color = 'red')+
  ggtitle('free SO2 in wine')
cat("summary of free sulfur dioxide\n")
summary(wine$free.sulfur.dioxide)
p2 <- ggplot(aes(x = bound.sulfur.dioxide),data = wine)+
  xlim(0,quantile(wine$bound.sulfur.dioxide,0.95))+
  geom_histogram(color = 'white',fill = 'blue')+
  geom_vline(aes(xintercept = median(bound.sulfur.dioxide)), color = 'red')+
  ggtitle('bound so2 in wine')
cat("summary of bound sulfur dioxide\n")
summary(wine$bound.sulfur.dioxide)
p3 <- ggplot(aes(x = total.sulfur.dioxide),data = wine)+
  xlim(0,quantile(wine$total.sulfur.dioxide,0.95))+
  geom_histogram(color = 'black',fill = 'white')+
  geom_vline(aes(xintercept = median(total.sulfur.dioxide)),color = 'red')+
  ggtitle('total so2 in wine')
cat("summary of total sulfur dioxide\n")
summary(wine$total.sulfur.dioxide)
p4 <- ggplot(aes(x = sulphates),data = wine)+
  geom_histogram(color = 'black',fill = 'white')+
  geom_vline(aes(xintercept = median(sulphates)),color = 'red')+
  ggtitle("sulphates in wine")
cat("summary of sulphates\n")
summary(wine$sulphates)
grid.arrange(p1,p2,p3,p4)
d1 <- ggplot(aes)
```

About 95% of the wine has the total sulfur dioxide that is well below the legal
limit. Bound sulfur dioxide is high in number comparing to the free sulfur
dioxide in wine. In other words the total sulfur dioxide contains more bound
SO2 than free SO2.

**density**
 
Density of the wine is closer to the density of water.According to winemakers 
sweet wine has more density than dry wine.

```{r echo=FALSE,message=FALSE,warning=FALSE,density}
ggplot(aes(density),data = wine)+
  geom_histogram(binwidth = 0.001,color = 'black',fill = 'white',
                 breaks = seq(0.985,1,0.001))+
  geom_vline(aes(xintercept = median(density)),color = 'red')
summary(wine$density)
```

Most of the wine has density between 0.985 to 1.000. In that about 50% data
falls in the range of 0.9917 to 0.9961

**quality**

Quality is the score between 0 and 10.

```{r echo=FALSE,message=FALSE,warning = FALSE}
ggplot(aes(x = quality),data = wine)+
  geom_histogram()
summary(wine$quality) 
```

There is a peak at wine. It means more number of the wine has the quality as 6.
Count is dropping after the quality 6. There is only 25% of the wine has more 
than 6 as a quality.

#univariate analysis

**What is the structure of your dataset?**

It has 4898 observations with 12 variables. In that 11 variables are of numeric
datatype and one is an interger.

**What are the main features of your dataset?**

Main feature of the  dataset is quality. All the other variables in the dataset
contributes to the quality oneway or another. Change in the one or more variable
may change the quality of the wine. So it is the main feature of the dataset.

**What other features in the dataset do you think will help your support your
investigation into your feature of your dataset?**

residual sugar,alcohol,fixed acidity,volatile acidity and citric acid contribute
aroma,taste and acidity to the wine so this will be important features to 
influence the quality of the white wine.

**Did you create any new variables in the dataset?**

Yes,I created two variables in the dataset. One is the taste_type. White wine 
bottles have labels to denote the sweetness.They are labeled based on the 
residual sugar content in the wine. So based on this i created a factor variable
to denote the sweet type.

Another varaibel is bound.sulfur.dioxide. When sulphates are added it provide
sulfur dioxide to the wine which inturn helps to fight againts the microbial
spoilage.

**of the features you investigated, were there any unusual distributions? Did 
you perform any operations on the data to tidy, adjust or change the form of the
data? If so,why did you do this? **

yes, density has the most unsual distribution.In density most of the wines fall
in the short range that leads to more wines fall in the same bin. So i have to 
set the binwidth, limits and the breaks on the x axis to see how it is 
distributed.

#Bivariate plot section.

**fixed.acidity by pH**

pH is the measure of the acidity of the solution. Higher the pH it contains less
acid and vice versa.

```{r echo=FALSE,message=FALSE,warning=FALSE,acid_vs_pH}
ggplot(aes(x = fixed.acidity,y = pH),data = wine)+
  geom_jitter(alpha = 1/5)+
  geom_smooth(method = 'lm')
cor(wine$fixed.acidity,wine$pH)
```

This show that pH and fixed acidity are negatively correlated.This prove the 
concept of measuring pH.They are correlated by -0.426. 

**chlorides vs quality **

Reason i choose to explore quality vs chloride is When the chloride content
becomes higher in the wine, it can affects the aroma of wine which in turn may
affects the quality.

```{r echo=FALSE,warning=FALSE,message= FALSE,chlorides_vs_quality}
ggplot(aes(x = chlorides,y = quality),data=subset(wine,wine$chlorides <= 0.1))+
  scale_x_continuous(limits = c(0,0.1),breaks = seq(0,0.1,0.01))+
  scale_y_continuous(limits = c(0,9),breaks = seq(0,9,1))+
  geom_vline(aes(xintercept = 0.06),color = 'red')+
  geom_jitter(alpha = 1/5)
cor(wine$chlorides,wine$quality)
```

In graph i use vertical line to denote the chlorides legal limit 0.06mg/L. In
this plot for the higher quality there is not much wines beyond the legal limit.
It shows that wine with high content of the chlorides don't have high quality 
and they have negative correlation which proves the point.But the strength of 
correlation is -0.21 that is not enough it affects the quality of the wine.

**alcohol vs residual.sugar**

During fermentation, sugars present in the wine are converted into alcohol.

```{r echo = FALSE,message = FALSE,warning = FALSE,alcohol_vs_residualsugar}
ggplot(aes(x = residual.sugar,y = alcohol),data = wine)+
  scale_x_continuous(limits = c(0,25),
                     breaks = seq(0,25,5))+
  geom_jitter(alpha = 1/5)+
  geom_smooth(method = 'lm',se = FALSE)
with(wine,cor(alcohol,residual.sugar))
```

alcohol and the residual sugars are negatively correlated. This show that the
higher the alcohol lesser the sugar content. There are correleted with -0.451.

**quality vs alcohol**

```{r echo = FALSE,warning = FALSE,message = FALSE,quality_vs_alcohol}
ggplot(aes(x=alcohol,y=quality),data=wine)+
  xlim(11,14)+
  geom_jitter(alpha=1/5)+
  geom_smooth(method='lm')
with(wine,cor(alcohol,quality))
```

Plot shows that quality and the alcohol are positively correlated with the 
correlation of 0.4355. There is a chance that wine with higheralcohol leads to
high quality wine but correlation does not imply causation , there may be other 
factors that influence this correlation.

**density vs alcohol **

when calculating the density of the wine , the contents of the denisty of the
wine are multiplied with their volume and added to give the wine density.
 
```{r echo = FALSE,warning = FALSE,message= FALSE,density_vs_alcohol}
ggplot(aes(x = alcohol,y = density),data=wine)+
    ylim(0.99,1)+
  geom_jitter(alpha=1/5)+
  geom_smooth(method='lm')
with(wine,cor(alcohol,density))
```

Density and alcohol are negatively correlated with -0.781. It shows that when
the alcohol content increases in wine the density of the wine tends to go down.

**density vs residual sugar**

```{r echo=FALSE,message =FALSE,warning=FALSE,density_vs_residualsugar}
ggplot(aes(x = residual.sugar,y = density),data=wine)+
  xlim(0,20)+
  ylim(0.985,1)+
  geom_point()+
  geom_smooth(method='lm')
with(wine,cor(residual.sugar,density))
```

There is a strong correlation between residual sugar and the density. It is 
a positively correlated with 0.8389. It shows that when residual sugar content 
in wine increases then the density of the wine also increases.

#Bivariate analysis

**Talk about some of the relationship you observed in this part of the 
investigation. How did the features of interest vary with the other features in 
the dataset?**

Quality is the important feature of the dataset. There is no interesting 
relationship for the quality in the dataset.When the chlorides increases it 
affects the wine quality as i assume but both are not strongly correlated to 
prove the assumption.When the alcohol content increases there is a increase in
the quality but the strength is not high enough between them to take into]
consideration.

**Did you observe any intersting relationship with the other features?**

Important relationship i found is the residual sugar with the alcohol and with
the density. During the fermantation sugars are converted into the alcohol.As 
alcohol is the product of the sugar, it results in the negative correlation 
between them. In the case of calculating density , density of the contents are 
considerd. Sugar has the higher density than the alcohol. When the wine has high
sugar content then it has a low alcohol content it automatically boost the 
density to high(sugar has higher density and sugar content high in wine). 
In other case when wine has high alcohol content then it has low sugar content
which in turn makes the density of the wine to low.

**What is the strongest realtionship you found?**

The strongest relationship is between the residual sugars and the density.
They are positively correlated with the strength of 0.8389.

#Multivariate Plot Section


```{r echo=FALSE,warning=FALSE,message=FALSE,mulitivariateplot1}
ggplot(aes(x = alcohol,y = residual.sugar , colour = taste_type),data = wine)+
  xlim(8,14)+
  geom_jitter(alpha=1/3)+
  facet_wrap(~quality,scales = 'free')+
  geom_smooth(method ='lm',color = 'red', se = FALSE)
by(wine$alcohol,wine$quality,summary)
```

In this plot you can see the cluster of the points shifting from left to right
as we go through the quality.It is clearly seen in the quality 5 and quality 7.
In quality 5 most of the points are in left side with alcohol range 8 to 12 
where as most of the points are in the right of the graph with the alcohol 
range 10 to 14. And you can also see the wine with the higher sugar content are
less in the high quality wine.

```{r echo=FALSE,message= FALSE,warning= FALSE,multivariateplot2}
ggplot(aes(x = total.sulfur.dioxide,y = chlorides),data = wine)+
  ylim(0,0.07)+
  xlim(50,200)+
  geom_jitter(aes(color = factor(quality)),alpha = 1/2)+
  geom_hline(yintercept = median(wine$chlorides))+
  geom_vline( xintercept = median(wine$total.sulfur.dioxide))+
  scale_color_brewer(type='qual')
conditions_met <- nrow(subset(wine,chlorides < median(wine$chlorides) & 
            total.sulfur.dioxide < median(total.sulfur.dioxide) & quality >= 7))
total_highquality_wines <- nrow(subset(wine,quality >= 7))
percentage <- (conditions_met / total_highquality_wines)*100 
cat("Percentage of the high quality wine with less 50% total chlorides 
    and total sulfur dioxide is")
percentage
```

when total sulfur dioxide and the chlorides are considered most of the high 
quality wine occupy the left bottom  of the vertical and horizantal  line.Nearly
53% high quality wines have less than 0.04 chlorides and 138.0 total sulfur 
dioxide.

#Multivariate Analysis

**Talk about some of the relationship you observed in this part of the 
investigation. Were there features that strengthened each other in terms of
look at your feature(s) of interest? **

Considering alcohol, residual sugars and the quality, Most of the high quality 
wine have high alcohol content and residual sugars are not playing a big part 
in deciding the quality. when chlorides and the total sulfur dioxide are
lower in wine it seems to strengthen the quality of wine.

**were there any interesting or surprising interections between the feature?**

Interesting Interaction is that about 53% of the high quality wine has the 
chloride and the total sulfur dioxide below their median values.

#Final plots and summary 

**Plot one**

```{r echo = FALSE,message=FALSE,warning=FALSE,plot1}
ggplot(aes(density),data = wine)+
  geom_histogram(binwidth = 0.001,color = 'black',fill = 'white',
                 breaks = seq(0.985,1,0.001))+
  geom_vline(aes(xintercept = median(density),color = "Median"))+
  labs(title = "density of the wine")+
  xlab('Density of wine g/cm^3')+
  ylab('Number of wines')+
  scale_color_manual("Legends",breaks = c("Median"),values = c("red"))
summary(wine$density)
```

**Description one**

About 75% of the wine has a density between 0.9871 to 0.9961 and in that 50% of 
wine is in between 0.9917 to 0.9967 that is 50% data fall within the distance of
0.005. Even though the density depands upon the content of the wine, it is 
interesting to see they fall with in the close range.

**Plot Two**

```{r echo=FALSE,message = FALSE, warning= FALSE,plot2}
legal_limit = 0.06
ggplot(aes(x = chlorides,y = quality),data=subset(wine,wine$chlorides <= 0.1))+
  scale_x_continuous(limits = c(0,0.1),breaks = seq(0,0.1,0.01))+
  scale_y_continuous(limits = c(0,9),breaks = seq(0,9,1))+
  geom_vline(aes(xintercept=legal_limit,colour="Legal Limit"))+
  geom_jitter(alpha = 1/5)+
  labs(title = "quality vs chlorides")+
  xlab("chlorides g/dm^3")+
  ylab("quality(score between 0 to 10)")+
  scale_colour_manual("Legands",breaks = c("Legal Limit"),values = c("Red"))
cor(wine$chlorides,wine$quality)
```

**Description Two **

Chlorides are due to the land where the vineyard is situated. Limit is
0.06 g/dm^3 which may be change in exceptional cases. But increase in the 
chlorides leads to change in the aroma of the wine which inturn affects the 
quality of the wine. In this most of high quality wine has chloride content 
well below the legal limit.

**Plot Three**

```{r echo=FALSE,warning=FALSE,message=FALSE}
  ggplot(aes(x = alcohol,y = residual.sugar,colour = taste_type),data = wine)+
  xlim(8,14)+
  geom_jitter(alpha=1/3)+
  facet_wrap(~quality,scales = 'free')+
  geom_smooth(method ='lm',color = 'red', se = FALSE)+
  ggtitle("residual sugar vs alcohol vs quality")+
  xlab("alcohol(% by volume)")+
  ylab("residual sugar (g/dm^3)")+
  scale_color_discrete(name = 'Taste Type')
by(wine$alcohol,wine$quality,summary)
```

**Description Three** 

For each quality the range of the alcohol which has more quality wines are 
varying.This can be clearly seen for quality 5 and quality 7. For quality 5 50%
wines have alcohol in the range of 9.200 to 10.30. For quality 7 about 50% wines
have alcohol are in 10.60 to 11.30.

#Reflection

I explored nearly 5000 wines in this dataset.what suprised me is that none of 
variables are strongly related with the quality. There are some strong relation
between density and sugar but these two are not correlated much with the 
quality. During the winemaking sugars are converted to alcohol, it is reflected
in this dataset. And the chlorides content may change the quality of the wine,
but there is not enough high quality wines to prove such a point.For residual 
sugars most of the wines have the low sugar content , thus it fail to show the 
relation with quality.For future investigation can add the variable like the 
date of the manufacturing of wine to see the ageing of wine has anything to do 
with the quality. wine with different chloride content and residual sugar 
content should be added in future. so that their influence on quality can be 
found. 




